---
title: "Importing and modifying WIC impact factors"
output: html_notebook
---

by Martin Brown, Martin.Brown@state.or.us

### Context and Introduction

Most WIC users will only be interested in WIC's "finished" impact factors, the ones that can be combined with solid waste data to estimate the life cycle impacts of waste streams -- as illustrated in *Example Application of the Waste Impact Calculator*.  Those finished impact factors are available in the "impactFactors/distributable" folder of the wic-base repository.

However, those finished impact factors have to come from somewhere.  Their source is modeling within the GaBi life cycle assessment system, as documented in *Impact modeling for the Waste Impact Calculator*.  But the factors exported by GaBi are not immediately usable in WIC.  They need to be imported and slightly modified, as described in *Technical Overview of the Waste Impact Calculator*, after which the resulting factors are saved for use elsewhere in WIC.  This R markdown document performs and explains that process. 

### Inputs and outputs

Inputs (stored within the "impactFactors/source" directory of the wic-base repository and/or zip files therein):

* several *.csv files with impact factor data
* gabiExportDate.txt
* impactCategoryPickerFixer.xlsx

Outputs (saved within the "distributable" directory of the impactFactors folder)

* impactFactors.csv
* impactFactors.RData

### Rationale for and outline of data processing

WIC's impact factors are defined within Peter Canepa's Gabi life cycle assessment software and then exported via Gabi's "pivot table export," which creates extremely large CSV files (as of this writing, ~90MB).  These should be found in the "source" folder of the impactFactors folder of the wic-base repository.

(Note: If the "source" folder in your downloaded repository contains only a zip file, it is because the uncompressed files are too large for github version control.  The code in this RMarkdown should unzip it, but if not, you can always unzip manually.)

Once the Gabi CSV files are imported, some filtering and relabeling of this voluminous data is necessary to get the impact factor down to a workable level of detail.  Following the method described in the *Technical Overview of the Waste Impact Calculator*, some of the impact factors are adjusted so they better match the life cycle stages relevant to the WIC model.  Finally, the filtered and modified impact factors are saved as data tables for use in other R sessions or use outside of WIC.  Those usable tables are saved in the "impactFactors/distributable" folder of the wic-base repo.

There are some complications along the way.  The most notable is the fact that there are actually 2 different files we need to import...

* a file with impact factors where greenhouse gas emissions calculations involving biogenic carbon include WIC's "slash" correction; and 
* a file with impact factors where greenhouse gas emissions involving biogenic carbon include WIC's "EPA FCS" correction. 

These biogenic factors need to be labeled appropriately and merged so there is only one impact factor file in the end.  See *Technical Overview of the Waste Impact Calculator* and *Impact modeling for the Waste Impact Calculator* for more detail about the substance of these corrections.

The following rough steps apply.

* Import the two Gabi "pivot table export" CSVs.
* Concatenate the two raw imports, labeling for slash and epaFcs.
* Filter and parse the raw imports.
* Filter and relabel impact categories
* Apply the Slash and EpaFCS labels to biogenic carbon fields.
* Save an 'unmodified' version of the file(s) for quality checking vs. Gabi model.
* Apply modifications called for in the *Technical Overview*.
* Save modified factors in suitable formats for future use.

### Setting up
```{r}
# check current working directory (should end in 
# "impactFactors" -- if not change filespecs appropriately)
getwd()
```
```{r}
# unzip the source files
unzip(
  zipfile="source/current/sourceData.zip", 
  overwrite = TRUE,
  exdir = "source/current"
  )
```

```{r}
# name Gabi export file filespecs
csSlashFileSpec  <-
  "source/current/wicFactorsFromGabiIncludingSlashMethod.csv"
csEpaFcsFileSpec <-
  "source/current/wicFactorsFromGabiIncludingEpaFcsMethod.csv"

# name the filespecs for two other files necessary for 
# the import
gabiExportDateFileSpec <-
  "source/current/gabiExportDate.txt"
impactCategoryPickerFixerFileSpec <-
  "source/current/impactCategoryPickerFixer.xlsx"

# misc parameters
randomFewNo <- 10  # useful for sizing diagnostic printouts
```

```{r}
# package version control:
# checkpoint() asks R to use R packages as of a certain date
# warning: the first time you use this it can take a while
library(checkpoint)
checkpoint(snapshot_date="2021-06-15")

# load R packages useful to this R markdown
library(tidyverse)
library(openxlsx)

```

### Importing the Gabi CSV's and concatenating them

Importing raw CSV's, labeling and concatenating them by fileSource, and printing out a few randomly chosen sample lines. 
```{r}
# import the raw CSV's
csSlashRaw <- 
  data.table::fread(
    file=csSlashFileSpec,
    sep=";",
    skip=10,     
#    nrows=200,  # use a row limit for testing
    header=TRUE
  ) %>%
  mutate(fileSource="Slash")

csEpaFcsRaw <- 
  data.table::fread(
    file=csEpaFcsFileSpec,
    sep=";",
    skip=10,     
#    nrows=200,  # use a row limit for testing
    header=TRUE
  ) %>%
  mutate(fileSource="EpaFcs")

# merge the CSV's (sources are now separable by fileSource)
csRaw <-
  bind_rows(csSlashRaw, csEpaFcsRaw)

# import the datestamp
dateStamps <- 
  read.csv(gabiExportDateFileSpec) %>%
  mutate(
    gabiExportDate=lubridate::ymd(gabiExportDate),
    wicImportDate=lubridate::today()
  )

# eliminate unnecessary objects
rm(csEpaFcsRaw, csSlashRaw)

# print out a few random records from the 
# imported records
csRaw %>% sample_n(randomFewNo) %>% print()

```


### Filtering and parsing the raw import.

As the randomly printed records above show, there's a lot of extraneous stuff in those Gabi imports.  I need to filter down to the rows that actually contain the rolled-up impact factors for individual materials and life cycle stages and/or dispositions, and then parse some of the values into more intuitive/distinct fields.

```{r}
csRawObjectPath <-
   csRaw %>% 
  # only use one particular Object path that represents 
  # summed up-impact factors
   filter(`Object path`=="Waste Impact Calculator <LC>") %>%
  # eliminate unneeded fields
   select(-Scenario, -`Object path`, 
         -`Object UUID`, -`Grouping: Nation`,
         -`Grouping: Type`, -`Grouping: Enterprise`, 
         - `Grouping: User Defined`,
         -`Inputs/Outputs single table view`
   ) %>%
   mutate(
     # parse the 'Full name/Local name' field
     # into material, LCstage, and disposition
      material=str_sub(
         `Full name/Local name`,
         start=1,
         end=str_locate(`Full name/Local name`,"_")[,1]-1
         ),
      theRestOfIt=str_sub(
         `Full name/Local name`,
         start=str_locate(`Full name/Local name`,"_")[,1]+1,
         end=str_length(`Full name/Local name`)
         ),
      LCstage=str_sub(
         theRestOfIt,
         start=1,
         end=str_locate(theRestOfIt,"_")[,1]-1
         ),
      disposition=str_sub(
         theRestOfIt,
         start=str_locate(theRestOfIt,"_")[,1]+1,
         end=str_length(theRestOfIt)
         )
   ) %>%
   select(-theRestOfIt)

# print a few random records from the data frame
# that was just created
csRawObjectPath %>% sample_n(randomFewNo) %>% print()
```

### Filter and relabel impact categories

The impactCategories that WIC needs are in the field Quantity, but they're all mixed up with corporate sources like "TRACI 2.1". Rather than writing a parsing routine, I'm going to refer to a lookup table that will do the parsing -- as well as choose a short list of impact categories to use and apply corrections to the names.

This lookup table is called "impactCategoryPickerFixer.xlsx" and should be found within the "source" folder in the current folder tree.  By all means use that file.  But if it can't be found, then the following code will create an Excel file in the necessary format:

```{r}
# get all the unique combinations of Quantity Folder, 
# Quantity, and Unit
impactCategoryPickerFixerStarter <-
  csRawObjectPath %>%
  select(c("Quantity Folder", "Quantity", "Unit")) %>%
  distinct() %>%
  # add columns I will need to fill in manually
  mutate(
    corporateSource = NA,
    impactCategoryLong = NA,
    shortList = as.logical(NA),
    impactCategory = NA,
    impactUnits = NA
  ) %>%
  arrange(`Quantity Folder`, Quantity)
# save it as an Excel file
write.xlsx(
  impactCategoryPickerFixerStarter,
  file="source/current/impactCategoryPickerFixerStarter.xlsx"
)
```

Assuming the lookup file can be found and is populated, now we import it and merge it with the impact factor information.  

```{r}
# import impactCategoryPickerFixer lookup file
impactCategoryPickerFixer <-
  readxl::read_excel(
    path=impactCategoryPickerFixerFileSpec
  )

# preview the join to see if there are any records
# in csRawObjectPath that WON'T match up with 
# the impactCategoryPickerFixer
csSelectedCheck <-
  anti_join(
    csRawObjectPath,
    impactCategoryPickerFixer,
    by=c("Quantity Folder", "Quantity")
  )

print(
  paste(
    "There are ", nrow(csSelectedCheck), 
    " row(s) that don't join.", sep=""
  )
)
```

If at this point, there are any rows that don't join, I should figure out the problem before moving on.  However, as of this writing, everything is getting a match.

Do the join:

```{r}
# merge with impact factor info from Gabi 
csSelected <-
  full_join(
    csRawObjectPath,
    impactCategoryPickerFixer,
    by=c("Quantity Folder", "Quantity")
  ) %>%
  # filter to keep only short-listed impact categories
  filter(shortList==TRUE) %>%
  # retain only certain fields, in a certain order
  select(fileSource, material, LCstage, disposition, 
         corporateSource, impactCategory, impactUnits,
         Value, impactCategoryLong) %>%
  # rename one field
  rename(impactFactor=Value) %>%
  # sort
  arrange(fileSource, material, LCstage, disposition,
          corporateSource, impactCategory, impactUnits,
          impactFactor) %>%
  # eliminate duplicate rows (though there shouldn't be any)
  distinct()

# print a few of the joined records 
csSelected %>% sample_n(randomFewNo) %>% print()

```

This is getting a lot closer to a usable impact factor file, but the LCstage and disposition factors still need a little recombining. Look at how they relate:

```{r}
table(csSelected$disposition, csSelected$LCstage, useNA = "ifany")
```

There are only two LCstages in the table above -- endOfLife and production.  Meanwhile the end-of-life dispositions include transport versions.  Let's assign the end-of-life transport stuff to its own LCstage.

Plus, othere's something weird going on with glass recycling, which has a disposition "recyclingToContainer <LC>". That's some sort of import typo and I will fix it.

```{r}
csSelected2 <-
   csSelected %>%
   mutate(
     # fix that weird Textiles typo
      # disposition = 
      #    ifelse(
      #       material=="Textiles" & disposition=="recyclingGeneric <LC>",
      #       "recyclingGeneric",
      #       disposition
      #    ),
     
      # fix a weird thing about glass recycling
      disposition =
        ifelse(
          disposition=="recyclingToContainer <LC>",
          "recyclingToContainer",
          disposition
        ),
      # create the endOfLifeTransport LCstage
      LCstage =
         ifelse(
           str_sub(
              string = disposition,
              start = str_length(disposition)-9,
              end = str_length(disposition)
           ) == "_transport" 
           & disposition != "production_transport",
           "endOfLifeTransport",
           LCstage
         ),
      # fix dispositions
      disposition =
        ifelse(
          str_sub(
              string = disposition,
              start = str_length(disposition)-9,
              end = str_length(disposition)
           ) == "_transport" 
           & disposition != "production_transport",
          str_sub(
            string=disposition,
            start=1,
            end=str_length(disposition)-10
          ),
          disposition
        )
   )
table(
   csSelected2$disposition, 
   csSelected2$LCstage,
   useNA = "ifany"
   )
```

Hmm it looks like something doesn't have a production_transport.
What is it?

```{r}
junk <-
   filter(
      csSelected2, 
      disposition=="production" | disposition=="production_transport"
   )
table(junk$material,junk$disposition)
```

It's food waste.  According to Peter Canepa, this is appropriate.

### Applying the Slash & EpaFcs labels 

Right now there's a lot of redundancy in this data table.  There are two versions of the impact factors in there, one where biogenic GHG factors were calculated using the Slash method, the other where they were calculated using the EpaFcs method.  Anything that is not a biogenic GHG factor is merely duplicated.  So I've got to correctly label those biogenic GHG categories, and remove the duplicates.

Here are the impactCategories currently in the table, and their relation to fileSource.

```{r}
table(csSelected2$impactCategory, csSelected2$fileSource, useNA="always")
```

So the impactCategories that include "biogenic" are the ones that need to be changed.

```{r}
# replace the generic qualifier "biogenic" in GHG-related
# impactCategory names with more specific qualifier
# drawn from the fileSource
csSelected3 <-
  csSelected2 %>%
  mutate(
    impactCategory =
        str_replace(
          impactCategory, 
          "(biogenic)",
          fileSource
        )
  )
table(
  csSelected3$impactCategory, csSelected3$fileSource, useNA="always"
)
```

That looks like it worked.  Now I can remove duplicates.

```{r}
# remove duplicates from impact factor collection
impactFactorsSelected <-
  csSelected3 %>%
  select(-fileSource) %>%
  distinct() %>%
  # sort as desired
  arrange(impactCategoryLong, material, disposition)
```

### Saving an "unaltered" import for checking.

At this point the impact factors from Gabi have been imported and filtered, but none of the quantities have been changed.  Let's save  the factors at this point so we can check the import itself against the original Gabi modelling files (if desired).

```{r}
write.xlsx(
  impactFactorsSelected,
  file="intermediate/importedButUnalteredForCheckingVsGabi.xlsx"
)
```

As of this writing, every check I have made on the import *process* has been good.  Questions about the values themselves I have created as "issues" in github.

### Modifying factors 

What I've got so far (in impactFactorsSelected) is good, but it's a little complex for use in most applications of the Waste Impact Calculator.  This set of factors has transport factors separated out for both production and end of life.  We don't need separate production and production transport factors.  

To do all this I'm going to have to split the impact factor file up and reassemble it after tweaking appropriately.

```{r}
# isolating production impact factors
ifProdJunk1 <- filter(
  impactFactorsSelected,
  LCstage == "production"
  )
# summing those up (combining production & production transport)
ifProdJunk2 <- summarise(
  group_by(
    ifProdJunk1, 
    material, LCstage, corporateSource, impactCategory, impactUnits,
    impactCategoryLong
  ),
  impactFactor=sum(impactFactor, na.rm=T)
  ) %>%
  ungroup() %>%
  mutate(disposition="production")
# so ifProdJunk2 has the production impact factors

# isolating end-of-life impact factors
ifEOLJunk1 <- filter(
  impactFactorsSelected,
  LCstage != "production"
  )

# now I can recombine the ifProdJunk2 & ifEOLJunk3
ifJunk1 <- bind_rows(
  ifProdJunk2,
  ifEOLJunk1
  )

```

Currently this list is still pretty long.
I'm going to cut a few dispositions as out of scope.
And I'm going to set the implied mileage of the end-of-life impact factors.

```{r}
# cutting some uncommon dispositions
ifJunk2 <- mutate(
  ifJunk1,
  cutFlag = case_when(
    disposition == "reuse" ~ TRUE
#    ,
#    disposition == "reuseContainer" ~ TRUE
    )
  ) %>%
  filter(is.na(cutFlag)) %>%
  select(-cutFlag) %>%
  # adding implied mileage for EOL transport impact factors
  mutate(
    impliedMiles = case_when(
      LCstage == "endOfLifeTransport" & 
        disposition != "recyclingToAggregate" ~ 180,
      LCstage == "endofLifeTransport" & 
        disposition == "recyclingToAggregate" ~ 20,
      TRUE ~ 180
      )
    ) %>%
  # reordering the variables
  select(material, LCstage, disposition, corporateSource,
         impactCategory,
         impactUnits, impactFactor, impliedMiles, impactCategoryLong)

# saving that as the dedicated impact factor file
impactFactors1 <- 
  ifJunk2 %>% 
  arrange(
    material, corporateSource, impactCategory, LCstage, disposition
  )
```

If it all looks good, I can delete the temporary objects.
```{r}
# deleting work objects
rm(ifEOLJunk1,  ifJunk1, ifJunk2, ifProdJunk1,
   ifProdJunk2,  impactFactorsSelected,
   csEpaFcsFileSpec, csRaw, csRawObjectPath,
   csSelected, csSelected2, csSelected3,
   csSlashFileSpec, impactCategoryPickerFixer,
   impactCategoryPickerFixerFileSpec,
   impactCategoryPickerFixerStarter,
   junk, csSelectedCheck)
```

The object impactFactors1 has most of the impact factors in a pretty clean, tidy state, but still needs a few more improvements.

Substantially, the end-of-life transport factor needs to be modified.  Canepa has calculated it so it includes end-of-life transport from the pickup point (e.g. curbside) to final processing.  However, for the purpose of using WIC to decide on local solid waste management this is not really the way to frame the transport.  Local managers are only in control of the first parts of that process... from the pickup point to the local disposal facility or the first recycling marketplace.  This logic is described more in the *Technical Overview* document.

So I need to split up the end-of-life transport factor for each material into two parts... the part managers can control, and those they can't.  

* Canepa says that the landfill distance issue is a good estimate of the controllable distance for each material.
* Anything beyond the landfill distance is really a part of the impacts for that end-of-life PROCESS (e.g. recycling).  So that should be added to the impacts for that process.

So, for every record in the impact factor database, I need to attach two things:

* The landfilling transport impact factor for that material; and
* The difference between the current impact factor and the landfilling transport impact factor

Then I need to apply one of those two things as corrections, but only to certain life cycle stages and dispositions.

```{r}
# getting landfill transport impacts
tempLandfillImpact <-
  impactFactors1 %>% 
  filter(disposition=="landfilling" & LCstage=="endOfLifeTransport") %>%
  select(-disposition, -impliedMiles, -LCstage) %>%
  rename(landfillTransportFactor = impactFactor)

# figuring the difference between the landfillTransportFactor
# for each material and the other transport impacts
tempTransportDifferences <-
  left_join(
    impactFactors1,
    tempLandfillImpact,
    by = c("material", "corporateSource", "impactCategory", "impactUnits", "impactCategoryLong")
  ) %>%
  filter(LCstage=="endOfLifeTransport") %>%
  select(-LCstage, -impliedMiles) %>%
  mutate(
    transportDifference = impactFactor - landfillTransportFactor
  ) %>%
  select(-impactFactor)

# attaching these two quantities to every row in the impact factor
# file
impactFactorsBetter <-
  left_join(
    impactFactors1,
    tempTransportDifferences,
    by = 
      c("material", "disposition", 
        "corporateSource", 
        "impactCategory", "impactUnits", 
        "impactCategoryLong")
  ) %>%
  # correcting when necessary
  mutate(
    correctFactorFlag =
      case_when(
        # no corrections for production impacts
        LCstage=="production"  ~ FALSE,
        str_sub(disposition, start=1, end=9) == "recycling" &
          disposition != "recyclingToAggregate" ~ TRUE,
        TRUE ~ FALSE
      ),
    impactFactor2 = 
      case_when(
        correctFactorFlag == FALSE ~ impactFactor,
        # recycling-related transport impacts set to landfill value
        correctFactorFlag == TRUE &
          LCstage == "endOfLifeTransport" ~ landfillTransportFactor,
        # recycling-related process impacts get an addition
        # due to non-local transport
        correctFactorFlag == TRUE &
          LCstage == "endOfLife" ~impactFactor + transportDifference
    )
  ) %>%
  # cleaning up the file
  select(
    -impactFactor, -landfillTransportFactor, -transportDifference,
    -correctFactorFlag
  ) %>%
  rename(impactFactor=impactFactor2)
```

In addition, I find that the material "nonrecyclables" doesn't have a production impact.   (This makes sense, as "nonrecyclables" aren't actually a material -- they're just a collection of miscellaneous things that can't be recycled.)  To fill in the database I'm just going to make an impact factor from the mean of the other materials.  However, other analysts down the line could do a better job by creating weighted average impact factors here based on an observed wastestream. 

```{r}
tempProductionImpacts <-
  impactFactorsBetter %>%
  filter(LCstage=="production") %>%
  select(-impliedMiles) %>%
  group_by(LCstage, disposition, impactCategory, impactUnits) %>%
  summarize(productionMean = mean(impactFactor))
impactFactorsBetter2 <-
  left_join(
    impactFactorsBetter,
    tempProductionImpacts,
    by = c("LCstage", "disposition", "impactCategory", "impactUnits")
  ) %>%
  mutate(
    impactFactor =
      ifelse(
        material=="Nonrecyclables" & LCstage=="production",
        productionMean,
        impactFactor
      )
  ) %>%
  select(-productionMean)
```

Finally, I need to address what to do with all of the miscellaneous materials that may turn up in a wastestream but are not specified by Peter C's impact factors.  What I'm going to do is come up with averages for each impact category and disposition, and then apply the material name "Other" to it.  As Peter C's list of impact factors grows, the materials covered by "Other" should shrink.

```{r}
tempOtherImpacts <-
  impactFactorsBetter2 %>%
  group_by(LCstage, disposition, corporateSource, 
           impactCategory, impactUnits, impactCategoryLong) %>%
  summarise(
    impactFactor=mean(impactFactor, na.rm=T),
    impliedMiles=mean(impliedMiles, na.rm=T)
  ) %>%
  ungroup() %>%
  mutate(material="Other") %>%
  filter(
    disposition %in% 
      c(
        "incinerationNoER",
        "landfilling",
        "production",
        "incinerationER",
        "recyclingGeneric"
      )
    )
impactFactorsBetter3 <-
  bind_rows(impactFactorsBetter2, tempOtherImpacts) %>%
  arrange(corporateSource, 
          impactCategory, material, LCstage, disposition
  )
```

Now I can save the file with a better name. Also need to add a date stamp here.

```{r}
# applying date stamp & saving with better name
impactFactorsDEQ <- 
  left_join(
    impactFactorsBetter3 %>% mutate(dummy=1),
    dateStamps %>% mutate(dummy=1),
    by="dummy"
  ) %>%
  select(-dummy)
# deleting temp objects
rm(impactFactorsBetter, impactFactorsBetter2, impactFactors1,
   tempTransportDifferences, tempLandfillImpact,
   tempProductionImpacts, tempOtherImpacts, impactFactorsBetter3)
# saving to disk files
saveRDS(
  impactFactorsDEQ, "distributable/impactFactors.RData"
)
write.csv(
  impactFactorsDEQ,
  "distributable/impactFactors.csv",
  row.names=FALSE
)
```

Ok, it looks like the factors have been imported.  There's more checking to do, but it looks pretty good.

I should print out a list of available materials and dispositions.

```{r}
availableMaterialsAndDispositions <-
  impactFactorsDEQ %>%
  select(material, disposition) %>%
  distinct() %>%
  mutate(available=TRUE)

availableMaterialsAndDispositionsFlipped <-
  pivot_wider(
    data = availableMaterialsAndDispositions,
    names_from = "disposition",
    values_from = available
  ) 

write.xlsx(
  availableMaterialsAndDispositionsFlipped,
  "distributable/availableMaterialsAndDispositions.xlsx"
)
```

