---
title: "Importing and modifying WIC impact factors"
output: html_notebook
---

by Martin Brown, Martin.Brown@state.or.us

### Context and Introduction

Most WIC users will only be interested in WIC's "finished" impact factors, the ones that can be combined with solid waste data to estimate the life cycle impacts of waste streams -- as illustrated in *Example Application of the Waste Impact Calculator*.  Those finished impact factors are available in the "distributable" folder of the current file tree.

However, those finished impact factors have to come from somewhere.  Their source is modeling within the GABI life cycle assessment system, as documented in *Impact modeling for the Waste Impact Calculator*.  But the factors exported by GABI are not immediately usable in WIC.  They need to be imported, filtered, and slightly modified, as described in *Technical Overview of the Waste Impact Calculator*, and the (slightly) transformed factors saved for use elsewhere in WIC.

This R markdown document performs and explains that process. 

### Inputs and outputs

Inputs (stored within the "source" directory of the current file tree)

* several *.csv files with impact factor data
* gabiExportDate.txt
* impactCategoryPickerFixer.xlsx

Outputs (saved within the "distributable" directory of the current file tree)

* impactFactors.csv
* impactFactors.RData

### Rationale for and outline of data processing

WIC's impact factors are defined within Peter Canepa's Gabi life cycle assessment software and then exported via Gabi's "pivot table export," which creates extremely large CSV files (as of this writing, ~80MB).  These should be found in the "source" folder of the current folder tree.

(Note: If the "source" folder is missing or empty, it may be because the Gabi export files are too large for github version control.  You may need to populate the "source" folder from zip files found elsewhere in the [wic-base release](https://github.com/OR-Dept-Environmental-Quality/wic-base/releases).)

Once the Gabi CSV files are imported, some filtering and relabeling of this voluminous data is necessary to get the impact factor down to a workable level of detail.  Following the method described in the *Technical Overview of the Waste Impact Calculator*, some of the impact factors are adjusted so they better match the life cycle stages relevant to the WIC model.  Finally, the filtered and modified impact factors are saved as data tables for use in other R sessions or use outside of WIC.  Those usable tables are saved the "distributable" folder within the current folder tree.

There are some complications along the way.  The most notable is the fact that there are actually 2 different files we need to import...

* impact factors where greenhouse gas emissions calculations involving biogenic carbon include WIC's "slash" correction; and 
* impact factors where greenhouse gas emissions involving biogenic carbon include WIC's "EPA FCS" correction. 

These biogenic factors need to be labeled appropriately and merged so there is only one impact factor file in the end.  See *Technical Overview of the Waste Impact Calculator* and *Impact modeling for the Waste Impact Calculator* for more detail about the substance of these corrections.

The following rough steps apply.

* Import the two Gabi "pivot table export" CSVs.
* Concatenate the two raw imports, labeling for slash and epaFcs.
* Filter and parse the raw imports.
* Filter and relabel impact categories
* Apply the Slash and EpaFCS labels to biogenic carbon fields.
* Save an 'unmodified' version of the file(s) for quality checking vs. Gabi model.
* Apply modifications called for in the *Technical Overview*.
* Save in suitable formats for future use.

### Setting up
```{r}
# check current working directory
getwd()

# load helpful R packages
library(tidyverse)
library(openxlsx)

# name Gabi "pivot table export" filespecs
csSlashFileSpec  <-
  "source/current/wicFactorsFromGabiIncludingSlashMethod.csv"
csEpaFcsFileSpec <-
  "source/current/wicFactorsFromGabiIncludingEpaFcsMethod.csv"

# name the filespecs for two other files necessary for 
# the import
gabiExportDateFileSpec <-
  "source/current/gabiExportDate.txt"
impactCategoryPickerFixerFileSpec <-
  "source/current/impactCategoryPickerFixer.xlsx"

# misc parameters
randomFewNo <- 10
```

### Importing the Gabi CSV's and concatenating them

Importing raw CSV's, labeling and concatenating them by fileSource, and printing out a few randomly chosen sample lines. 
```{r}
# import the raw CSV's
csSlashRaw <- 
  data.table::fread(
    file=csSlashFileSpec,
    sep=";",
    skip=10,     
#    nrows=200,  # use a row limit for testing
    header=TRUE
  ) %>%
  mutate(fileSource="Slash")

csEpaFcsRaw <- 
  data.table::fread(
    file=csEpaFcsFileSpec,
    sep=";",
    skip=10,     
#    nrows=200,  # use a row limit for testing
    header=TRUE
  ) %>%
  mutate(fileSource="EpaFcs")

# merge the CSV's (sources are now separable by fileSource)
csRaw <-
  bind_rows(csSlashRaw, csEpaFcsRaw)

# import the datestamp
dateStamps <- 
  read.csv(gabiExportDateFileSpec) %>%
  mutate(
    gabiExportDate=lubridate::ymd(gabiExportDate),
    wicImportDate=lubridate::today()
  )

# eliminate unnecessary objects
rm(csEpaFcsRaw, csSlashRaw)

# print out a few random records from the 
# imported records
csRaw %>% sample_n(randomFewNo) %>% print()

```


### Filtering and parsing the raw import.

As those random records  show, there's a lot of extraneous stuff in those Gabi imports.  I need to filter down to the rows that actually contain the rolled-up impact factors, and then parse some of the values into more intuitive/distinct fields.

```{r}
csRawObjectPath <-
   csRaw %>% 
  # only use one particular Object path that represents 
  # summed up impact factors
   filter(`Object path`=="Waste Impact Calculator <LC>") %>%
  # eliminate unneeded fields
   select(-Scenario, -`Object path`, 
         -`Object UUID`, -`Grouping: Nation`,
         -`Grouping: Type`, -`Grouping: Enterprise`, 
         - `Grouping: User Defined`,
         -`Inputs/Outputs single table view`
   ) %>%
   mutate(
     # parse the 'Full name/Local name' field
     # into material, LCstage, and disposition
      material=str_sub(
         `Full name/Local name`,
         start=1,
         end=str_locate(`Full name/Local name`,"_")[,1]-1
         ),
      theRestOfIt=str_sub(
         `Full name/Local name`,
         start=str_locate(`Full name/Local name`,"_")[,1]+1,
         end=str_length(`Full name/Local name`)
         ),
      LCstage=str_sub(
         theRestOfIt,
         start=1,
         end=str_locate(theRestOfIt,"_")[,1]-1
         ),
      disposition=str_sub(
         theRestOfIt,
         start=str_locate(theRestOfIt,"_")[,1]+1,
         end=str_length(theRestOfIt)
         )
   ) %>%
   select(-theRestOfIt)

# print the random records
csRawObjectPath %>% sample_n(randomFewNo) %>% print()
```

### Filter and relabel impact categories

The impactCategories that WIC needs are in the field Quantity, but they're all mixed up with corporate sources like "TRACI 2.1". Rather than writing a parsing routine, I'm going to refer to a lookup table that will do the parsing -- as well as choose a short list of impact categories to use and apply corrections to the names.

First I need to create the lookup table -- something I should only do once, as the table is meant to be edited manually.  But here is the format:

```{r}
# get all the unique combinations of Quantity Folder, Quantity, and Unit
impactCategoryPickerFixerStarter <-
  csRawObjectPath %>%
  select(c("Quantity Folder", "Quantity", "Unit")) %>%
  distinct() %>%
  # add columns I will need to fill in manually
  mutate(
    corporateSource = NA,
    impactCategoryLong = NA,
    shortList = as.logical(NA),
    impactCategory = NA,
    impactUnits = NA
  ) %>%
  arrange(`Quantity Folder`, Quantity)
# saving it as an Excel file
write.xlsx(
  impactCategoryPickerFixerStarter,
  file="source/current/impactCategoryPickerFixerStarter.xlsx"
)
```

Ok, I created the table and filled it in, but saved it with a different filename just in case.

Let's import the lookup file and then merge it ... hopefully we won't lose any records.  Once it's joined, I can start carving out only the fields and impact categories I want.

```{r}
# importing impactCategoryPickerFixer lookup file
impactCategoryPickerFixer <-
  readxl::read_excel(
    path=impactCategoryPickerFixerFileSpec
  )

# merging 
csSelected <-
  full_join(
    csRawObjectPath,
    impactCategoryPickerFixer,
    by=c("Quantity Folder", "Quantity")
  ) %>%
  # filtering only short-listed impact categories
  filter(shortList==TRUE) %>%
  # keep only fields I want in this order
  select(fileSource, material, LCstage, disposition, 
         corporateSource, impactCategory, impactUnits,
         Value, impactCategoryLong) %>%
  # rename one field
  rename(impactFactor=Value) %>%
  # sort
  arrange(fileSource, material, LCstage, disposition,
          corporateSource, impactCategory, impactUnits,
          impactFactor) %>%
  # eliminate duplicates (though there shouldn't be any)
  distinct()

# print a few at random
csSelected %>% sample_n(10) %>% print()

```

This is getting a lot closer to a usable impact factor file, but the LCstage and disposition factors still need a little recombining. Look at how they relate:

```{r}
table(csSelected$disposition, csSelected$LCstage, useNA = "ifany")
```

There are only two LCstages -- endOfLife and production.  Meanwhile the end-of-life dispositions include transport versions.  Let's assign the end-of-life transport stuff to its own LCstage.

Plus there's something weird going on with Textiles, which has a disposition called "recyclingGeneric <LC>".  

```{r}
csSelected2 <-
   csSelected %>%
   mutate(
     # fix that weird Textiles thing
      disposition = 
         ifelse(
            material=="Textiles" & disposition=="recyclingGeneric <LC>",
            "recyclingGeneric",
            disposition
         ),
      # create the endOfLifeTransport LCstage
      LCstage =
         ifelse(
           str_sub(
              string = disposition,
              start = str_length(disposition)-9,
              end = str_length(disposition)
           ) == "_transport" 
           & disposition != "production_transport",
           "endOfLifeTransport",
           LCstage
         ),
      # fix dispositions
      disposition =
        ifelse(
          str_sub(
              string = disposition,
              start = str_length(disposition)-9,
              end = str_length(disposition)
           ) == "_transport" 
           & disposition != "production_transport",
          str_sub(
            string=disposition,
            start=1,
            end=str_length(disposition)-10
          ),
          disposition
        )
   )
table(
   csSelected2$disposition, 
   csSelected2$LCstage,
   useNA = "ifany"
   )
```

Hmm it looks like something doesn't have a production_transport.
What is it?

```{r}
junk <-
   filter(
      csSelected2, 
      disposition=="production" | disposition=="production_transport"
   )
table(junk$material,junk$disposition)
```

It's food waste.  Check that later.

### Applying the Slash & EpaFcs labels 

Right now there's a lot of redundancy in this data table.  There are two versions of the impact factors in there, one where biogenic GHG factors were calculated using the Slash method, the other where they were calculated using the EpaFcs method.  Anything that is not a biogenic GHG factor is merely duplicated.  So I've got to correctly label those biogenic GHG categories, and remove the duplicates.

Here are the impactCategories currently in the table, and their relation to fileSource.

```{r}
table(csSelected2$impactCategory, csSelected2$fileSource, useNA="always")
```

So the impactCategories that include "biogenic" are the ones that need to be changed.

```{r}
csSelected3 <-
  csSelected2 %>%
  mutate(
    impactCategory =
        str_replace(
          impactCategory, 
          "(biogenic)",
          fileSource
        )
  )
table(
  csSelected3$impactCategory, csSelected3$fileSource, useNA="always"
)
```

That looks like it worked.  Now I can remove duplicates.

```{r}
impactFactorsSelected <-
  csSelected3 %>%
  select(-fileSource) %>%
  distinct() %>%
  arrange(impactCategoryLong, material, disposition)
```

### Saving an "unaltered" import for checking.

```{r}
write.xlsx(
  impactFactorsSelected,
  file="source/current/importedButUnalteredForCheckingVsGabi.xlsx"
)
```

Hmm... everything so far has checked out.  I have q's about the substance of the impact factors, but the import process has been ok.
Need to move on to modifying the factors.

### Modifying factors 

What I've got so far (in impactFactorsSelected) is good, but it's a little complex for use in most applications of the Waste Impact Calculator.  This set of factors has transport factors separated out for both production and end of life.  We don't need separate production and production transport factors.  Moreover the names for the dispositions are really long.  

To do all this I'm going to have to split the impact factor file up and reassemble it after tweaking appropriately.

```{r}
# isolating production impact factors
ifProdJunk1 <- filter(
  impactFactorsSelected %>% select(-impactCategoryLong),
  LCstage == "production"
  )
# summing those up (combining production & production transport)
ifProdJunk2 <- summarise(
  group_by(
    ifProdJunk1, 
    material, LCstage, corporateSource, impactCategory, impactUnits
  ),
  impactFactor=sum(impactFactor, na.rm=T)
  ) %>%
  ungroup() %>%
  mutate(disposition="production")
# so ifProdJunk2 has the production impact factors

# isolating end-of-life impact factors
ifEOLJunk1 <- filter(
  impactFactorsSelected %>% select(-impactCategoryLong),
  LCstage != "production"
  )

# now I can recombine the ifProdJunk2 & ifEOLJunk3
ifJunk1 <- bind_rows(
  ifProdJunk2,
  ifEOLJunk1
  )

```

Currently this list is still pretty long.
I'm going to cut some dispositions as unrealistic.
I'm going to rename some dispositions for clarity.
And I'm going to set the implied mileage of the end-of-life impact factors.

```{r}
# cutting some uncommon dispositions
ifJunk2 <- mutate(
  ifJunk1,
  cutFlag = case_when(
#    disposition == "incinerationNoER" ~ TRUE,
#    disposition == "recyclingFiberglass" ~ TRUE,
    disposition == "reuse" ~ TRUE,
    disposition == "reuseContainer" ~ TRUE
    )
  ) %>%
  filter(is.na(cutFlag)) %>%
  select(-cutFlag) %>%
  # renaming some awkward dispositions
  mutate(
    disposition = case_when(
      disposition == "recyclingGeneric" ~ "recycling",
      disposition == "recyclingAggregate" ~ "useAsAggregate",
      disposition == "recyclingToAggregate" ~ "useAsAggregate",
      disposition == "recyclingToContainer" ~ "recycling",
      disposition == "incinerationER" ~ "combustion",
      disposition == "incinerationNoER" ~ "combustionNoER",
      TRUE ~ disposition
      )
    ) %>%
  # adding implied mileage for EOL transport impact factors
  mutate(
    impliedMiles = case_when(
      LCstage == "endOfLifeTransport" & disposition == "landfilling" ~ 180,
      LCstage == "endOfLifeTransport" & disposition == "recycling" ~ 180,
      LCstage == "endOfLifeTransport" & disposition == "composting" ~ 180,
      LCstage == "endOfLifeTransport" & disposition == "useAsAggregate" ~ 20,
      LCstage == "endOfLifeTransport" & disposition == "anaerobicDigestion" ~ 180,
      LCstage == "endOfLifeTransport" & disposition == "combustion" ~ 180,
      TRUE ~ 180
      )
    ) %>%
  # reordering the variables
  select(material, LCstage, disposition, corporateSource,
         impactCategory,
         impactUnits, impactFactor, impliedMiles)

# saving that as the dedicated impact factor file
impactFactors1 <- 
  ifJunk2 %>% 
  arrange(
    material, corporateSource, impactCategory, LCstage, disposition
  )
```

If it all looks good, I can delete the temporary objects.
```{r}
# deleting work objects
rm(ifEOLJunk1,  ifJunk1, ifJunk2, ifProdJunk1,
   ifProdJunk2,  impactFactorsSelected,
   csEpaFcsFileSpec, csRaw, csRawObjectPath,
   csSelected, csSelected2, csSelected3,
   csSlashFileSpec, impactCategoryPickerFixer,
   impactCategoryPickerFixerFileSpec,
   impactCategoryPickerFixerStarter,
   junk)
```

The object impactFactors1 has most of the impact factors in a pretty clean, tidy state, but still needs a few more improvements.

Substantially, the end-of-life transport factor needs to be modified.  Canepa has calculated it so it includes end-of-life transport from the pickup point (e.g. curbside) to final processing.  However, for the purpose of using WIC to decide on local solid waste management this is not really the way to frame the transport.  Local managers are only in control of the first parts of that process... from the pickup point to the local disposal facility or the first recycling marketplace.

So I need to split up the end-of-life transport factor for each material into two parts... the part managers can control, and those they can't.  

* Canepa says that the landfill distance issum a good estimate of the controllable distance for each material.
* Anything beyond the landfill distance is really a part of the impacts for that end-of-life PROCESS (e.g. recycling).  So that should be added to the impacts for that process.

So, for every record in the impact factor database, I need to attach two things:

* The landfilling transport impact factor for that material; and
* The difference between the current impact factor and the landfilling transport impact factor

Then I need to apply one of those two things as corrections, but only to certain life cycle stages and dispositions.

```{r}
# getting landfill transport impacts
tempLandfillImpact <-
  impactFactors1 %>% 
  filter(disposition=="landfilling" & LCstage=="endOfLifeTransport") %>%
  select(-disposition, -impliedMiles, -LCstage) %>%
  rename(landfillTransportFactor = impactFactor)

# figuring the difference between the landfillTransportFactor
# for each material and the other transport impacts
tempTransportDifferences <-
  left_join(
    impactFactors1,
    tempLandfillImpact,
    by = c("material", "corporateSource", "impactCategory", "impactUnits")
  ) %>%
  filter(LCstage=="endOfLifeTransport") %>%
  select(-LCstage, -impliedMiles) %>%
  mutate(
    transportDifference = impactFactor - landfillTransportFactor
  ) %>%
  select(-impactFactor)

# attaching these two quantities to every row in the impact factor
# file
impactFactorsBetter <-
  left_join(
    impactFactors1,
    tempTransportDifferences,
    by = 
      c("material", "disposition", 
        "corporateSource", 
        "impactCategory", "impactUnits")
  ) %>%
  # correcting when necessary
  mutate(
    impactFactor2 = case_when(
      # no correction for production impacts
      LCstage == "production" ~ impactFactor,
      # no correction for certain end-of-life impacts
      (LCstage == "endOfLifeTransport" | LCstage == "endOfLife") & 
         (disposition == "landfilling" | 
         disposition == "combustion" |
         disposition == "combustionNoER" |
         disposition == "anaerobicDigestion" | 
         disposition == "composting" |
         disposition == "useAsAggregate") 
          ~ impactFactor,
      # recycling-related transport impacts set to landfill value
        LCstage == "endOfLifeTransport" &
        (disposition != "landfilling" & 
         disposition != "combustion" &
         disposition != "combustionNoER" & 
         disposition != "anaerobicDigestion" & 
         disposition != "composting" &
         disposition != "useAsAggregate") 
          ~ landfillTransportFactor,
      # recycling-related process impacts get an addition
      # due to non-local transport
        LCstage == "endOfLife" &
        (disposition != "landfilling" & 
         disposition != "combustion" &
         disposition != "combustionNoER" &
         disposition != "anaerobicDigestion" & 
         disposition != "composting" &
         disposition != "useAsAggregate") 
          ~ impactFactor + transportDifference
    )
  ) %>%
  # cleaning up the file
  select(
    -impactFactor, -landfillTransportFactor, -transportDifference
  ) %>%
  rename(impactFactor=impactFactor2)
```

In addition, I find that the material "nonrecyclables" doesn't a production impact.  I'm just going to take it from the mean of the other materials.  This is probably something I should improve later, though, as it isn't weighted by any real waste stream.

```{r}
tempProductionImpacts <-
  impactFactorsBetter %>%
  filter(LCstage=="production") %>%
  select(-impliedMiles) %>%
  group_by(LCstage, disposition, impactCategory, impactUnits) %>%
  summarize(productionMean = mean(impactFactor))
impactFactorsBetter2 <-
  left_join(
    impactFactorsBetter,
    tempProductionImpacts,
    by = c("LCstage", "disposition", "impactCategory", "impactUnits")
  ) %>%
  mutate(
    impactFactor =
      ifelse(
        material=="Nonrecyclables" & LCstage=="production",
        productionMean,
        impactFactor
      )
  ) %>%
  select(-productionMean)
```

Finally, I need to address what to do with all of the miscellaneous materials that may turn up in a wastestream but are not specified by Peter C's impact factors.  What I'm going to do is come up with averages for each impact category and disposition, and then apply the material name "Other" to it.  As Peter C's list of impact factors grows, the materials covered by "Other" should shrink.

```{r}
tempOtherImpacts <-
  impactFactorsBetter2 %>%
  group_by(LCstage, disposition, corporateSource, 
           impactCategory, impactUnits) %>%
  summarise(
    impactFactor=mean(impactFactor, na.rm=T),
    impliedMiles=mean(impliedMiles, na.rm=T)
  ) %>%
  ungroup() %>%
  mutate(material="Other")
impactFactorsBetter3 <-
  bind_rows(impactFactorsBetter2, tempOtherImpacts) %>%
  arrange(corporateSource, 
          impactCategory, material, LCstage, disposition
  )
```

Now I can save the file with a better name. Also need to add a date stamp here.

```{r}
# applying date stamp & saving with better name
impactFactorsDEQ <- 
  left_join(
    impactFactorsBetter3 %>% mutate(dummy=1),
    dateStamps %>% mutate(dummy=1),
    by="dummy"
  ) %>%
  select(-dummy)
# deleting temp objects
rm(impactFactorsBetter, impactFactorsBetter2, impactFactors1,
   tempTransportDifferences, tempLandfillImpact,
   tempProductionImpacts, tempOtherImpacts, impactFactorsBetter3)
# saving to disk files
saveRDS(
  impactFactorsDEQ, "distributable/impactFactors.RData"
)
write.csv(
  impactFactorsDEQ,
  "distributable/impactFactors.csv",
  row.names=FALSE
)
```

Ok, it looks like the factors have been imported.  There's more checking to do, but it looks pretty good.